Temperature and moisture effects on greenhouse gas emissions from deep active-layer

### 2016-04-22 manuscript/cpcrw_manuscript.Rmd
New draft- discussion fleshed out, figure legends

@@

High latitude ecosystems are being subjected to rapid changes in climate {IPCC, 2013 #3916} and fire occurrence and intensity {Kasischke, 2010 #2631}, notably in northwestern North America and Alaska {Hinzman, 2005 #4315}{Ju, 2016 #4320}. This will have a wide variety of effects on both ecosystems and society: in particular, rising temperatures and increasing fire will likely result in permafrost degradation {Pastick, 2015 #4311}{Zhang, 2015 #4372} and changes in soil temperature, with subsequent hydrology changes that will influence soil greenhouse gas (GHG) fluxes to the atmosphere. Such fluxes are a large component of the global C cycle and could thus promote a positive climate feedback {Treat, 2015 #4217}{Koven, 2011 #3092}.

@@

Temperature and moisture typically have strong and often interactive influences on soil GHG emissions. A long literature of laboratory incubations, field observations, and meta-analyses has documented increased CO~2~ fluxes with rising temperature {Davidson, 2006 #1669}{Hashimoto, 2015 #4231}{Treat, 2015 #4217}. GHG responses to wetting and thawing dynamics are much less certain, with substantial variability between studies {Kim, 2012 #2863}. The anaerobic conditions common following permafrost thaw are expected to lower CO~2~ emissions but increase those of CH~4~ {Treat, 2015 #4217;Treat, 2014 #4213}; such interactions are critical to examine in the course of long-term incubation experiments {Elberling, 2013 #3720}. More uniformly across the landscape, however, decadal warming, and perhaps drying, trends in Alaska {Bieniek, 2014 #4389} will work to counteract these effects, with uncertain consequences for the carbon cycle and climate system.

@@

**Importance and context**

Rises in boreal air temperatures, and unpredictable precipitation changes, will warm and dry many soils, increase vegetation stress {Ju, 2016 #4320}, degrade permafrost and deepen the active layer {Schuur, 2015 #4319}. The resulting changes in the soil carbon cycle may already be visible {Bond-Lamberty, 2012 #3480} but are highly uncertain going forward, increasing the need for controlled laboratory studies of GHG emission dynamics from high-latitude soils.

**Previous results**

@@

More generally, in a pan-Arctic synthesis of anaerobic soil incubations, Treat et al. {, 2015 #4217} reported mean CO~2~ rates of 47 (all mineral soils) and 101 (for 20-100 cm soils) µg C-CO~2~ g C^-1^ day^-1^, somewhat lower than our aerobic incubation results. Weiss et al. {, 2015 #4317} found CO~2~ production from Yedoma permafrost samples to be correlated with both percent C and N, consistent with our results (**Table 1**). TODO.

**Temperature versus moisture sensitivity**

@@

In contrast, the temperature sensitivities observed in this experiment are low (all less than 2.0, even when controlling for changes in soil moisture). Such values are not unprecedented in comparison to a wide range of other laboratory soil incubations {Hamdi, 2013 #3664}. It is important to note, however, that even if increased temperature alone may not always drive C mineralization rates in forest-mineral soils {Giardina, 2000 #2159}, but it is frequently linked with increases in soil moisture content can lead to changes in microbial community structure, amplifying GHG fluxes (Xue et al, 2016).  TODO: PEYTON - WHAT IS THIS CITATION?

**Methane**

We observed very low, but consistently positive, CH~4~ production from these unsaturated mineral soils. This is contrast to many field studies, e.g. Schaufler et al. {, 2010 #4391} who found consistent CH~4~ uptake (oxidation) in dry conditions at a boreal site.

Treat et al. found a median CO~2~:CH~4~ production ratios of 387 for boreal sites, far lower than (but consistent with) our observed ratios of several million.

**Limitations and weaknesses**

Incubation experiments can only go so far though...

Laboratory soil incubations have weaknesses, for example in what can be inferred about temperature sensitivity {Podrebarac, 2016 #4365}{Hamdi, 2013 #3664}, but their controlled environments provide an important way to elucidate the key mechanisms controlling GHG from high-latitude soils {Schuur, 2015 #4319}.

these are not surface layer soils (where majority of microbial activity and C mineralization of labile C takes place) and that taking them out of depth (where they are not exposed to O2 as readily, etc.) for our laboratory incubation may differ from in situ CO2/CH4 production.

Necessity of integrated lab-field-modeling experiments...

[//]: (comment comment comment)

@@

**Table 1.** Linear mixed-effects model parameters, testing effects of gravimetric water content (unitless), temperature (°C), soil C (%), soil N (%),  and dissolved organic carbon (mg kg^-1^) on individual core CO~2~ fluxes (+`r FLUX_ADDITION` µg C g C^-1^ day^-1^). Dependent variable has units of log(µg C g C^-1^ day^-1^). Columns include parameter value; standard error (SE); degrees of freedom (DF); T statistic; and P value.

@@

**Table 2.** Linear mixed-effects model parameters, testing effects of gravimetric water content (unitless), temperature (°C), soil C (%), soil N (%),  and dissolved organic carbon (mg kg^-1^) on log-transformed, individual core CH~4~ fluxes (+`r FLUX_ADDITION` µg C g C^-1^ day^-1^). Dependent variable has units of log(µg C g C^-1^ day^-1^). Columns include parameter value; standard error (SE); degrees of freedom (DF); T statistic; and P value.

@@

**Figure 2.** A faceted (Temperature ~ Treatment) plot showing normalized CO2 flux means and errors by incubation time.

@@

**Figure 3.** A faceted (Temperature ~ Treatment) plot showing normalized CH4 flux means and errors by incubation time.

@@

**Figure 4.** Cumulative fluxes over the incubation.

>>>>

[//]: This is the Rmarkdown file

@@

High latitude ecosystems are being subjected to rapid changes in climate {IPCC, 2013 #3916} and fire occurrence and intensity {Kasischke, 2010 #2631}, notably in northwestern North America and Alaska {Hinzman, 2005 #4315}{Ju, 2016 #4320}. This will have a wide variety of effects on both ecosystems and society: in particular, rising temperatures and increasing fire will likely result in permafrost degradation {Pastick, 2015 #4311}{Zhang, 2015 #4372} and changes in soil temperature, with subsequent hydrology changes that will influence soil greenhouse gas (GHG) fluxes to the atmosphere. Such fluxes are a large component of the global C cycle and, because of the high C stocks of northern soils {Tarnocai, 2009 #2231}, could result in a significant and positive climate feedback {Treat, 2015 #4217}{Koven, 2011 #3092}.

@@

Temperature and moisture typically have strong and often interactive influences on soil GHG emissions. Laboratory incubations, field observations, and meta-analyses have documented increased CO~2~, and under some conditions CH~4~ {Olefeldt, 2013 #3456}, fluxes with rising temperature {Davidson, 2006 #1669}{Hashimoto, 2015 #4231}{Treat, 2015 #4217}. GHG responses to wetting and thawing dynamics are much less certain, with substantial variability between studies {Kim, 2012 #2863}. The anaerobic conditions common following permafrost thaw are expected to lower CO~2~ emissions but increase those of CH~4~ {Treat, 2015 #4217;Treat, 2014 #4213}, and such interactions are critical to examine in the course of long-term incubation experiments {Elberling, 2013 #3720}. More uniformly across the landscape, however, decadal warming, and perhaps drying, trends in Alaska {Bieniek, 2014 #4389} will work to counteract these effects, with uncertain consequences for the carbon cycle and climate system.

@@

@@

More generally, in a pan-Arctic synthesis of anaerobic soil incubations, Treat et al. {, 2015 #4217} reported mean CO~2~ rates of 47 (all mineral soils) and 101 (for 20-100 cm soils) µg C-CO~2~ g C^-1^ day^-1^, somewhat lower than our aerobic incubation results. Weiss et al. {, 2015 #4317} found CO~2~ production from Yedoma permafrost samples to be correlated with both percent C and N, consistent with our results (**Table 1**). The response of soil biota to stresses such as drought tend to differ between soil types, but be broadly similar across biomes and climatic conditions {Manzoni, 2012 #3545}.

*Temperature versus moisture sensitivity*

@@

In contrast, the temperature sensitivities observed in this experiment were low (all less than 2.0, even when controlling for changes in soil moisture). Such values are not unprecedented in comparison to a wide range of other laboratory soil incubations {Hamdi, 2013 #3664}. It is important to note, however, that even if increased temperature alone may not always drive C mineralization rates in forest-mineral soils {Giardina, 2000 #2159}, but it is frequently linked with increases in soil moisture content can lead to changes in microbial community structure and GHG fluxes {Xue, 2016 #4361}.

*Methane*

We observed very low, but consistently positive, CH~4~ production from these unsaturated mineral soils. This is contrast to many field studies, e.g. Schaufler et al. {, 2010 #4391} who found consistent CH~4~ uptake (oxidation) in dry conditions at a boreal site. Our results are however broadly consistent with data from 65 studies summarized by Olefeldt et al. {Olefeldt, 2013 #3456}, who found that CH~4~ emissions were more sensitive to soil temperature in wetter ecosystems; it would have been a surprise if the little methanogenic activity in our upland, well-drained soils was temperature-sensitive at all.

Methane was thus a far smaller C flux than CO~2~ from these soils. This is true more generally: for example, Treat et al. found a median CO~2~:CH~4~ production ratios of 387 for boreal sites, far lower than (but consistent with) our observed ratios of several million. MORE.

*Limitations and weaknesses*

There were weaknesses in our approach and experimental design that should be considered. Laboratory experiments offer precise control, but lack the *in situ* nature of field manipulations {Sistla, 2013 #3666}, raising uncertainties to what degree their results can be extrapolated. They also have more specific weaknesses, for example in what can be inferred about temperature sensitivity {Podrebarac, 2016 #4365}{Hamdi, 2013 #3664}. Nonetheless, the controlled environments of incubations provide an important way to elucidate the key mechanisms controlling GHG from high-latitude soils {Schuur, 2015 #4319}.

We focused on a experimental drought, rather than flooding, because of the well-drained nature of the field site: it is unlikely that the mid-slope forest we sampled in will ever suffer from thermokarst or excessive soil moisture, but too-dry conditions are a serious possibility in this low-precipitation ecosystem {Barber, 2000 #1320}. In addition, the soils here are not surface layer soils (where the majority of microbial activity and C mineralization of labile C takes place); taking them out of depth (where they are less exposed to O~2~, for example) may significantly change the abiotic conditions to which the microbial community is adapted. MORE?

*Conclusions*

Rises in boreal air temperatures, and unpredictable precipitation changes, will warm and dry many soils, increase vegetation stress {Ju, 2016 #4320}{Barber, 2000 #1320}, degrade permafrost and deepen the active layer {Schuur, 2015 #4319}. The resulting changes in the soil carbon cycle may already be visible {Bond-Lamberty, 2012 #3480} but are highly uncertain going forward, increasing the need for controlled laboratory studies, ideally integrated with field and modeling experiments, of GHG emission dynamics from high-latitude soils.

@@

**Table 1.** Linear mixed-effects model parameters, testing effects of temperature (°C), gravimetric water content (unitless), soil C (%), soil N (%),  and dissolved organic carbon (mg kg^-1^) on individual core CO~2~ fluxes (+`r FLUX_ADDITION` µg C g C^-1^ day^-1^); a colon (":") indicates an interaction. Dependent variable has units of log(µg C g C^-1^ day^-1^). Columns include parameter value; standard error (SE); degrees of freedom (DF); T statistic; and P value.

@@

**Table 2.** Linear mixed-effects model parameters, testing effects of temperature (°C), gravimetric water content (unitless), soil N (%),and dissolved organic carbon (DOC, mg kg^-1^) on log-transformed, individual core CH~4~ fluxes (+`r FLUX_ADDITION` µg C g C^-1^ day^-1^); a colon (":") indicates an interaction. Dependent variable has units of log(µg C g C^-1^ day^-1^). Columns include parameter value; standard error (SE); degrees of freedom (DF); T statistic; and P value.

@@

**Figure 2.** Mass-normalized CO~2~ fluxes over the 100-day incubation, by temperature (4 and 20 °C, rows) and treatment (field moisture, drought, and controlled drought; columns). Error bars show core-to-core standard deviation.

@@

**Figure 3.** Mass-normalized CH~4~ fluxes over the 100-day incubation, by temperature (4 and 20 °C, rows) and treatment (field moisture, drought, and controlled drought; columns). Error bars show core-to-core standard deviation.

@@

**Figure 4.** Cumulative C fluxes (mg) over the incubation, by gas (CO~2~ and CH~4~, rows) and treatment (columns). TODO: these need to be normalized.
